---
title: "workflow"
author: "Thijs Janzen"
date: "8/12/2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{workflow}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
library(ggplot2)
library(magrittr)
library(tidyr)
library(nodeSub)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 6)
knitr::opts_chunk$set(fig.height = 6)
```

# NodeSub workflow

Given a tree, we can generate a NodeSubstitution alignment as follows (please note that for this example, we use an extremely short sequence length to reduce simulation time. Such short sequence lengths might not be appropriate for more serious applications.)

```{r create alignment}
seq_length <- 100
sub_rate <- 1 / seq_length

input_tree <- TreeSim::sim.bd.taxa(n = 100,
                                   numbsim = 1,
                                   lambda = 1,
                                   mu = 0.1,
                                   complete = TRUE)[[1]]

target_alignment <- sim_unlinked(phy = input_tree,
                                 rate1 = sub_rate,
                                 rate2 = sub_rate,
                                 l = seq_length,
                                 node_time = 0.3)
```

We can then proceed to create a Twin alignment (e.g. an alignment with the exact same number of accumulated substitutions, given the same tree, but using a 'normal' substitution model)

```{r twin alignment}
comp_alignment <- create_equal_alignment(
                        input_tree = geiger::drop.extinct(input_tree),
                        # can only work on trees without extinct branches
                        sub_rate = sub_rate,
                        alignment_result = target_alignment)
```
Now we have two alignments, one generated using the Node Substitution model, and one using standard strict clock model. For both we can perform phylogenetic inference to get the resulting tree. The aim is to compare the posterior distribution of trees with the true tree that we started with 'input_tree', and estimate the error invoked by the node substitution model.

```{r infer_phylogeny}

if (beastier::is_beast2_installed()) {

  node_posterior <- infer_phylogeny(target_alignment$alignment,
                                    "node_posterior",
                                    clock_prior = beautier::create_strict_clock_model(clock_rate_param =    beautier::create_clock_rate_param(value = sub_rate)),
                                    chain_length = 1e5,
                                    burnin = 0.1,
                                    working_dir = getwd(),
                                    sub_rate = sub_rate)

  reference_posterior <- infer_phylogeny(comp_alignment$alignment,
                                         "reference_posterior",
                                         burnin = 0.1,
                                         clock_prior = beautier::create_strict_clock_model(clock_rate_param = beautier::create_clock_rate_param(value = comp_alignment$adjusted_rate)),
                                         chain_length = 1e5,
                                         working_dir = getwd,
                                         sub_rate = sub_rate)
}
```

Having two posterior distributions of trees, we can compare them based on 
summary statistics (note 09-14-2023: the beta statistic is no longer supported,
as the calculating package was removed from CRAN).

```{r stats}
if (beastier::is_beast2_installed()) {
node_stats <- calc_sum_stats(node_posterior$all_trees,
                             input_tree)
ref_stats  <- calc_sum_stats(reference_posterior$all_trees,
                            input_tree)

node_stats$differences$method <- "node_sub"
ref_stats$differences$method  <- "reference"

results <- rbind(node_stats$differences,
                 ref_stats$differences)
}
```

We can now visualize the differences. 
```{r plot results}
if (beastier::is_beast2_installed()) {
  results %>%
  tidyr::gather(key = "statistic", value = "val", -method) %>%
  dplyr::filter(statistic != "num_tips") %>%
  ggplot(aes(x = method, y = val)) +
    geom_boxplot() +
    facet_wrap(~statistic, scales = "free")
}
```

We find that the beta, gamma and nLTT statistic yield very similar results. 
However, the JSD, crown age and mean branch length statistics indicate that when
using the node substitution model, errors arise, where the crown age of the trees
inferred from the node substitution alignment is too high, resulting in a higher
mean branch length and a difference in the JSD statistic.
